Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-04-09 main.tex
Added text to the intro based on my draft steer control paper.

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}
\usepackage{todonotes}

@@

\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
Subjected to Mechanical Perturbations at Low Speeds}

@@

\section{Introduction}
%
Single actor bicycle crashes are associated with a large number of unreported
and reported injuries and even sometimes death. At low speeds (less than 15
km/h or so) bicycles are no longer self stable and become more difficult to
control for the rider. Low speed crashes may be reduced if the bicycle was self
stable. Since at least the 1980s it known that applying a steering torque
proportional to the roll rate can stabilize a single track vehicle at very low
speeds. If the robotic control of steering can stabilize a bicycle, it may
reduce the control required by the rider to successfully manage balancing
tasks. We have developed such a balance assisting bicycle and desire to
evaluate whether it helps the rider in situations they are likely to fall.

@@

Cluster-mean centering shows there to be no variation between participants,
i.e. that rider skill (the only participant differentiating factor) is not a
predictor. We expected there to be a variation between participants in how well
they are able to resist the perturbation. However, this was not true. This
allowed us to use a simpler single-level logistic regression model, instead of
a multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll & steer angles at the time of perturbation. We
also include interaction effects between the balance assist state and the other
four variables.

@@

\bibliographystyle{plain}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\bibliography{references}

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage{natbib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\title{Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds
When Subjected to Mechanical Perturbations}

@@

\section{Introduction}
%
Single actor bicycle crashes are associated with a large number of unreported
and reported injuries. At low speeds (less than 15 km/h or so) bicycles
themselves are not self-stable and become more difficult to control for the
rider. Low speed crashes may be reduced if the bicycle was self stable at these
speeds. Since at least the 1980s it known that applying a steering torque
proportional to the roll rate can stabilize a single track vehicle at very low
speeds. If the robotic control of steering can stabilize a bicycle, it may
reduce the control required by the rider to successfully manage balancing
tasks. We have developed such a balance assisting bicycle and desire to
evaluate whether it helps the rider in situations they are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined throughout the 1970s.
\Citet{vanZytveld1975} seems to be the first to attempt to robotically
stabilize a small motorbike with a controlled inverted pendulum that mimicked
rider lean, but was not successful in demonstrating what his control model
predicted. In his control model, he recognized that feedback of the vehicle
roll angle and angular rate was essential to stabilize the vehicle. It was not
until the mid 1980s that someone successfully demonstrated an automatically
balanced motorcycle~\cite{Ruijs1986a} and they did so with a steer motor. Ruijs
and Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
show how the gains need to change with respect to vehicle speed. Roll motion
feedback is the simplest controller that can stabilize a single track vehicle
above a minimum speed when you are not concerned with wobble destablization.
But Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability reduces falling when ridden.  Most of these robotic bicycles and
motorcycles designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel. The effect of this automatic
control gives the ability to change the dynamics, to a degree, of the human
controlled plant.

The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. It is ideal for showing the effects of a roll motion feedback
driven steer motor on the dynamics. The linear model can be described by the
fourth order state space equations:

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
\end{align}

The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
populated with expressions that are functions of 27 geometric and inertial
parameters of the nonholonomic mulitbody system made up of four rigid bodies:
two wheels, front frame, and rear frame.

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:

\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} - \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right] \mathbf{B}_\delta\right) \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left[\mathbf{A} - \left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \mathbf{B}_\delta\right]\) have negative real parts for
\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
the speed range where the bicycle is stable can be maximized within the limits
of any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller.

In this paper, we test whether a gain scheduled steer motor controlled bicycle
that is riderless stable at a low speed range is beneficial in preventing
whether the rider will fall. To do this, we apply varying magintidue mechanical
perturbations to the handlebars while the rider is balancing the bicycle on a
treadmill and then assess the riders' probability of falling with the balance
assist system on versus off.

@@

Cluster-mean centering shows there to be no variation between participants,
i.e. that rider skill (the only participant differentiating factor) is not a
predictor. We expected there to be a variation between participants in how well
they are able to resist the perturbation. However, this was not true. This
allowed us to use a simpler single-level logistic regression model, instead of
a multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation. We
also include interaction effects between the balance assist state and the other
four variables.

@@

\bibliographystyle{apalike}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\bibliography{references}
